# Introduction {#intro}

You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(intro). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(methods).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

### geocoding
 - geocoding example with tmaptools using open street map

```{r geocode-osm example}
# uses OSM
library(tmap)
library(tmaptools)
library(dplyr)
tex_cap <-tmaptools::geocode_OSM("Texas Capital", 
        as.sf = TRUE) %>% 
  glimpse()
```

### interactive mapping

```{r mapview example}
library(mapview)
mapview(tex_cap)
```

### Choropleth map
The package maps (automatically installed and loaded with ggplot2) provides maps of the USA, with state and county borders, that can be retrieved and converted as sf objects:
```{r tmap example}
library(sf)
library(maps)
counties <- st_as_sf(map("county", plot = FALSE, fill = TRUE))
counties <- subset(counties, grepl("texas", counties$ID) & !grepl('missouri,texas',counties$ID))
counties$area <- as.numeric(st_area(counties))
head(counties)

tm_shape(counties) +
  tm_polygons("area", 
              style="quantile", 
              title="Texas Counties Area")
```