# Reading in Spatial Data

- There are several ways we typically get spatial data into R:
  + Load spatial files we have on our machine or from remote source
  + Load spatial data that is part of an R package
  + Grab data using API (often making use of particular R packages)
  + Converting flat files with x,y data to spatial data
  + Geocoding data (we saw example of this at beginning)


For reading and writing vector and raster data in R, the three primary packages you'll use are:

 * `sf` or `rgdal` for vector formats such as [ESRI Shapefiles](https://en.wikipedia.org/wiki/Shapefile),  [GeoJSON](https://en.wikipedia.org/wiki/GeoJSON), and [GPX](https://en.wikipedia.org/wiki/GPS_Exchange_Format) - both packages use **OGR**, which is a library under the [GDAL](https://en.wikipedia.org/wiki/GDAL) source tree,under the hood
 * `raster` for raster formats such as [GeoTIFF](https://en.wikipedia.org/wiki/GeoTIFF) or [ESRI or ASCII grid](https://en.wikipedia.org/wiki/Esri_grid) using [GDAL](https://en.wikipedia.org/wiki/GDAL) under the hood
 
We can quickly discover supported I/O vector formats either via `sf` or `rgdal`:
```{r message=FALSE, warning=FALSE, error=FALSE}
library(knitr)
library(sf)
library(rgdal)
kable(head(ogrDrivers(),n=5))
kable(head(st_drivers(what='vector'),n=5))
```

As well as I/O raster formats via `sf`:
```{r message=FALSE, warning=FALSE, error=FALSE}
library(knitr)
kable(head(st_drivers(what='raster'),n=5))
```