# Reading in Spatial Data

- There are several ways we typically get spatial data into R:
  + Load spatial files we have on our machine or from remote source
  + Load spatial data that is part of an R package
  + Grab data using API (often making use of particular R packages)
  + Converting flat files with x,y data to spatial data
  + Geocoding data (we saw example of this at beginning)


For reading and writing vector and raster data in R, the three primary packages you'll use are:

 * `sf` or `rgdal` for vector formats such as [ESRI Shapefiles](https://en.wikipedia.org/wiki/Shapefile),  [GeoJSON](https://en.wikipedia.org/wiki/GeoJSON), and [GPX](https://en.wikipedia.org/wiki/GPS_Exchange_Format) - both packages use **OGR**, which is a library under the [GDAL](https://en.wikipedia.org/wiki/GDAL) source tree,under the hood
 * `raster` for raster formats such as [GeoTIFF](https://en.wikipedia.org/wiki/GeoTIFF) or [ESRI or ASCII grid](https://en.wikipedia.org/wiki/Esri_grid) using [GDAL](https://en.wikipedia.org/wiki/GDAL) under the hood
 
We can quickly discover supported I/O vector formats either via `sf` or `rgdal`:
```{r message=FALSE, warning=FALSE, error=FALSE}
library(knitr)
library(sf)
library(rgdal)
print(paste0('There are ',st_drivers("vector") %>% nrow(), ' vector drivers available using st_read or read_sf'))
kable(head(ogrDrivers(),n=5))
kable(head(st_drivers(what='vector'),n=5))
```

As well as I/O raster formats via `sf`:
```{r message=FALSE, warning=FALSE, error=FALSE}
library(knitr)
print(paste0('There are ',st_drivers(what='raster') %>% nrow(), ' raster drivers available'))
kable(head(st_drivers(what='raster'),n=5))
```

## Reading in vector data
`sf` can be used to read numerous file types:

- Shapefiles
- Geodatabases
- Geopackages
- Geojson
- Spatial database files

### Shapefiles
Typically working with vector GIS data we work with ESRI shapefiles or geodatabases - here we have an example of how one would read in a shapefile using `sf`:
```{r shapefile_read, message=FALSE, warning=FALSE, error=FALSE}
# download.file("ftp://ftp.gis.oregon.gov/adminbound/citylim_2017.zip","citylim_2017.zip")
# unzip("citylim_2017.zip", exdir = ".")
library(sf)
citylims <- st_read("citylim_2017.shp")
options(scipen=3)
plot(citylims$geometry, axes=T, main='Oregon City Limits') # plot it!
```

### `st_read` versus `read_sf` {.challenge}
Above, I didn't pass any parameters to st_read - typically I would pass the parameters `quiet=TRUE` and `stringsAsFactors=FALSE` - why would this be a good practice in general?

### Answer  {.solution}
`read_sf` is an `sf` alternative to st_read (see [this](https://keen-swartz-3146c4.netlify.com/intro.html#read) section 1.2.2). Try reading in `citylims` data above using `read_sf` and notice difference, and check out `help(read_sf)`.  `read_sf` and `write`sf` are simply aliases for st_read and st_write with modified default arguments.  Big differences are:

- stringsAsFactors=FALSE
- quiet=TRUE
- as_tibble=TRUE

### Geodatabases
We use `st_read` or `read_sf` similarly for reading in an ESRI file geodatabase feature:
```{r geodatabase_read, message=FALSE, warning=FALSE, error=FALSE}
# download.file("ftp://ftp.gis.oregon.gov/adminbound/OregonStateParks_20181010.zip", "OregonStateParks.zip")
# unzip("OregonStateParks.zip", exdir = ".")
library(ggplot2)
fgdb = "OregonStateParks_20181010.gdb"

# List all feature classes in a file geodatabase
st_layers(fgdb)

# Read the feature class
parks <- st_read(dsn=fgdb,layer="LO_PARKS")
ggplot(parks) + geom_sf()
```

### Geopackages
Another spatial file format is the geopackage. Let's try a quick read and write of geopackage data.  First we'll read in a geopackage using data that comes with `sf` using `dplyr` syntax just to show something a bit different and use `read_sf` as an alternative to `st_read`.  You may want to try writing the data back out as a `geopackage` as well. 

Quick question: What are a couple advantages of `geopackages` over `shapefiles`?
```{r geopackage_read, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
nc <- system.file("gpkg/nc.gpkg", package="sf") %>% read_sf() # reads in
glimpse(nc)
```

### Open spatial data sources
There is a wealth of open spatial data accessible online now via static URLs or APIs - a few examples include [Data.gov](https://catalog.data.gov/dataset?metadata_type=geospatial), NASA [SECAC Portal](http://sedac.ciesin.columbia.edu/), [Natural Earth](http://www.naturalearthdata.com/), [UNEP GEOdata](http://geodata.grid.unep.ch/), and countless others listed here at [Free GIS Data](https://freegisdata.rtwilson.com/)

### Spatial data from R packages
There are also a number of R packages written specifically to provide access to geospatial data - below are a few and we'll step through some examples of pulling in data using some of these packages.
```{r datapackages, echo=FALSE}
datapackages = tibble::tribble(
  ~`Package name`, ~Description,
  "USABoundaries", "Provide historic and contemporary boundaries of the US",
  "tigris", "Download and use US Census TIGER/Line Shapefiles in R",
  "tidycensus", "Uses Census American Community API to return tidyverse and optionally sf ready data frames",
  "FedData","Functions for downloading geospatial data from several federal sources",
  "elevatr", "Access elevation data from various APIs (by Jeff Hollister)",
  "getlandsat", "Provides access to Landsat 8 data.",
  "osmdata", "Download and import of OpenStreetMap data.",
  "raster", "The `getData()` function downloads and imports administrative country, SRTM/ASTER elevation, WorldClim data.",
  "rnaturalearth", "Functions to download Natural Earth vector and raster data, including world country borders.",
  "rnoaa", "An R interface to National Oceanic and Atmospheric Administration (NOAA) climate data.",
  "rWBclimate", "An access to the World Bank climate data."
)
knitr::kable(datapackages, caption = "Example R packages for spatial data retrieval.")
```

Below is an example of pulling in US states using the `rnaturalearth` package - note that the default is to pull in data as `sp` objects and we coerce to `sf`.  Also take a look at the chained operation using dplyr.  Try changing the filter or a parameter in ggplot.
```{r naturalearth, , message=FALSE, warning=FALSE, error=FALSE}
library(rnaturalearth)
library(dplyr)
library(ggplot2)
states <- ne_states(country = 'United States of America')
states_sf <- st_as_sf(states)
states_sf %>%
  dplyr::filter(!name %in% c('Hawaii','Alaska') & !is.na(name)) %>%
  ggplot + geom_sf()
```

### Read in raster data
Here we use the `getData` function in the `raster` package to download elevation into a `RasterLayer` and grab administrative boundaries from a database of global administrative boundaries - warning: sometimes `getData` function has trouble accessing the server and download can be a bit slow. Here we see as well how we can use vector spataio polygon data to crop raster data.
```{r rasterdata, message=FALSE, warning=FALSE, error=FALSE}
library(raster)
US <- getData("GADM",country="USA",level=2)
Benton <- US[US$NAME_1=='Oregon' & US$NAME_2=='Benton',]
elev <- getData('SRTM', lon=-123, lat=44)
elev <- crop(elev, Benton)
elev <- mask(elev, Benton)
plot(Benton, main="Elevation (m) in Benton County, Oregon", axes=T)
plot(elev, add=TRUE)
plot(Benton, add=TRUE)
```

### Read in OpenStreetMap data
The [osmdata](https://github.com/ropensci/osmdata) package is a fantastic resource for leveraging the OpenStreetMap (OSM) database.
```{r osmdata , message=FALSE, warning=FALSE, error=FALSE}
library(osmdata)
library(mapview)

footway <- opq(bbox = "corvallis oregon") %>% 
  add_osm_feature(key = "highway", value = "footway") %>% 
  osmdata_sf()
footway <- footway$osm_lines
mapview(footway$geometry)
```


```{r osmdata2 , message=FALSE, warning=FALSE, error=FALSE}
library(osmdata)
library(mapview)
rstrnts <- opq(bbox = "corvallis oregon") %>% 
    add_osm_feature(key = "amenity", value = "restaurant") %>%
    osmdata_sf()
rstrnts <- rstrnts$osm_points

mapview(rstrnts$geometry)
```

We often have flat files, locally on our machine or accessed elsewhere, that have coordinate information which we would like to make spatial:
```{r vector_drivers, message=FALSE, warning=FALSE, error=FALSE}
library(devtools)
library(readr)
library(ggplot2)
# install_github("mhweber/awra2020spatial", force=TRUE)
library(awra2020spatial)
gages = read_csv(system.file("extdata/Gages_flowdata.csv", package = "awra2020spatial"))

gages_sf <- gages %>%
  st_as_sf(coords = c("LON_SITE", "LAT_SITE"), crs = 4269, remove = FALSE) %>%
  dplyr::select(STATION_NM,LON_SITE, LAT_SITE)

ggplot() + geom_sf(data=gages_sf)
```


