# Geoprocessing
## Lesson Goals

A quick look at a couple typical topological operations (spatial subsetting, spatial joins, dissolve) using `sf`

## Example one
### Spatial Subsetting
Let's look at the bike paths and parks data in the `awra2020spatial` package. A typical spatial question we might ask of our data is 'what trails go through parks in town?' A great feature of `sf` is it supports spatial indexing:
```{r subset, message=FALSE, warning=FALSE, error=FALSE}
library(sf)
library(awra2020spatial)
data(parks)
data(bike_paths)

plot(bike_paths$geoms, col='green', axes=T)
plot(parks$geoms, col='blue', add=T)
paths_in_parks <- bike_paths[parks,]
plot(paths_in_parks$geoms, col='red', lwd = 2, add=T)
title(main='Bike Paths in Parks in Austin')
```

## Example two
### Spatial Join
First we'll use chained operations to select just a couple columns from both bike_paths and parks, and then we'll do a spatial join operation in `sf`.  Note again, when we do a select on just attribute column, the geometry column remains - geometry is sticky in `sf`!
```{r join, message=FALSE, warning=FALSE, error=FALSE}
library(dplyr)
bike_paths <- bike_paths %>% 
  dplyr::select(ROUTE_NAME)
parks <- parks %>% 
  dplyr::select(LOCATION_NAME, ZIPCODE,PARK_TYPE)
parks_bike_paths <- st_join(parks, bike_paths) # st_intersects is the default
glimpse(parks_bike_paths)
```

## Example Three
### Dissolve
We can perform a spatial dissolve in `sf` using `dplyr` `group_by` and `summarize` functions with an `sf` object!
Note that we could pull down `tidycensus` at tract level, but instead we want to look at running a dissolve to get from block group to tract level
```{r dissolve, message=FALSE, warning=FALSE, error=FALSE}
library(ggplot2)
parks$AREA <- st_area(parks)
parks_zip <- parks %>% 
  group_by(ZIPCODE) %>%
  summarise(AREA = sum(AREA)) %>%
  ggplot() + geom_sf(aes(fill=(ZIPCODE))) +
  ggtitle("Austin Parks by Zip Code") + 
  theme_bw()
parks_zip
```